Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-07-02 thirdPaper/draft/typeConeAssociahedra.tex
removed comments

\smallskip
\begin{itemize}
\item \textbf{Gentle associahedra}\yann{Ref physicien accordions ?} were constructed in~\cite{PaluPilaudPlamondon-nonkissing} as polytopal realizations of finite non-kissing complexes of gentle bound quivers. These complexes encompass two families of simplicial complexes studied independently in the literature: on the one hand the grid associahedra introduced by T.~K.~Petersen, P.~Pylyavskyy and D.~Speyer in~\cite{PetersenPylyavskyySpeyer} for a staircase shape, studied by F.~Santos, C.~Stump and V.~Welker~\cite{SantosStumpWelker} for rectangular shapes, and extended by T.~McConville in~\cite{McConville} for arbitrary grid shapes; and on the other hand the Stokes polytopes and accordion associahedra studied by Y.~Baryshnikov~\cite{Baryshnikov}, F.~Chapoton~\cite{Chapoton-quadrangulations}, A.~Garver and T.~McConville~\cite{GarverMcConville} and T.~Manneville and V.~Pilaud~\cite{MannevillePilaud-accordion}. The latter, sometimes called accordiohedra, have also~recently appeared in the mathematical physics litterature concerning scattering amplitudes~\cite{BanerjeeLaddhaRaman-StokesPolytopes, Raman-PositiveGeometry}, with realizations inspired by those in~\cite{ArkaniHamedBaiHeYan}. It was shown in~\cite{PaluPilaudPlamondon-nonkissing, BrustleDouvilleMousavandThomasYildirim} that the non-kissing complex of a gentle bound quiver provides a combinatorial model for the support $\tau$-tilting complex~\cite{AdachiIyamaReiten} of the associated gentle algebra. Even for finite non-kissing complexes, it turns out that the type cone of the non-kissing fan is not always simplicial (see \cref{exm:nonSimplicialTypeConeNKC}). We prove however that the type cone is always simplicial when the gentle bound quiver is brick (at least $2$ relations in any cycle) and $2$-acyclic (no oriented cycle of length~$2$). Moreover, we precisely describe the facet-defining inequalities of the type cone in terms of linear dependences of $\b{g}$-vectors involved in certain mesh relations. We thus automatically derive a construction similar to that of~\cite{ArkaniHamedBaiHeYan, BazierMatteDouvilleMousavandThomasYildirim} describing all polytopal realizations of the $\b{g}$-vector fans of these brick and $2$-acyclic gentle bound quivers as sections of a positive orthant by well-chosen affine spaces (see \cref{subsec:typeConeCA}). Again, this result mainly relies on finding which relations minimally generate the relations between the $\b{g}$-vectors in the non-kissing complex. We present both a purely combinatorial proof (see \cref{subsec:typeConeCA}) and a purely algebraic proof for (non-necessarily gentle) brick algebras making use of extriangulated categories (see \cref{sec:extricats}).
Our algebraic proof is based on an analogue of a result of M.~Auslander~\cite{Auslander1984} relating to minimal generators of Grothendieck groups.
Along the way, we need to generalize, for extriangulated categories, several results from cluster-tilting theory on indices~\cite{DehyKeller} and on abelian quotients~\cite{BuanMarshReiten, KellerReiten} that were known for triangulated categories. We note, however, that our assumptions are not quite the expected ones. This is related to the fact that the Grothendieck groups of cluster categories are badly behaved (\eg the Grothendieck group of a cluster category of type~$A_2$ is trivial). So as to overcome this difficulty, we replace the triangulated structure of cluster categories by some relative extriangulated structure, before considering Grothendieck groups. This explains why we consider projective objects rather than cluster-tilting objects for extriangulated categories.

@@

\medskip\yann{Ref physicien accordions ?}
To conclude, it is our hope that the present paper will participate to the interactions between combinatorial geometry, representation theory, and mathematical physics; in particular through further developments of the geometric approach to scattering amplitudes that has flourished during the last years~\cite{ArkaniHamedTrnka-Amplituhedron,ArkaniHamedBourjailyCachazoGoncharovPostnikovTrnka-GrassmannianGeometryScatteringAmplitudes}. The kinematic associahedron from~\cite{ArkaniHamedBaiHeYan} is one of the most recent ``positive geometries''~\cite{ArkaniHamedBaiLam-PositiveGeometries} that has emerged in the study of scattering amplitudes.

@@

\cref{prop::KRextricat} below extends similar results from~\cite{BuanMarshReiten,KellerReiten,KoenigZhu,IyamaYoshino}\yann{Add reference silting Iyama} (see also \cref{prop::functor-F}) to the setting under consideration. We note that the proof requires minor modifications.

>>>>

\smallskip
\begin{itemize}
\item \textbf{Gentle associahedra} were constructed in~\cite{PaluPilaudPlamondon-nonkissing} as polytopal realizations of finite non-kissing complexes of gentle bound quivers. These complexes encompass two families of simplicial complexes studied independently in the literature: on the one hand the grid associahedra introduced by T.~K.~Petersen, P.~Pylyavskyy and D.~Speyer in~\cite{PetersenPylyavskyySpeyer} for a staircase shape, studied by F.~Santos, C.~Stump and V.~Welker~\cite{SantosStumpWelker} for rectangular shapes, and extended by T.~McConville in~\cite{McConville} for arbitrary grid shapes; and on the other hand the Stokes polytopes and accordion associahedra studied by Y.~Baryshnikov~\cite{Baryshnikov}, F.~Chapoton~\cite{Chapoton-quadrangulations}, A.~Garver and T.~McConville~\cite{GarverMcConville} and T.~Manneville and V.~Pilaud~\cite{MannevillePilaud-accordion}. The latter, sometimes called accordiohedra, have also~recently appeared in the mathematical physics litterature concerning scattering amplitudes~\cite{BanerjeeLaddhaRaman-StokesPolytopes, Raman-PositiveGeometry}, with realizations inspired by those in~\cite{ArkaniHamedBaiHeYan}. It was shown in~\cite{PaluPilaudPlamondon-nonkissing, BrustleDouvilleMousavandThomasYildirim} that the non-kissing complex of a gentle bound quiver provides a combinatorial model for the support $\tau$-tilting complex~\cite{AdachiIyamaReiten} of the associated gentle algebra. Even for finite non-kissing complexes, it turns out that the type cone of the non-kissing fan is not always simplicial (see \cref{exm:nonSimplicialTypeConeNKC}). We prove however that the type cone is always simplicial when the gentle bound quiver is brick (at least $2$ relations in any cycle) and $2$-acyclic (no oriented cycle of length~$2$). Moreover, we precisely describe the facet-defining inequalities of the type cone in terms of linear dependences of $\b{g}$-vectors involved in certain mesh relations. We thus automatically derive a construction similar to that of~\cite{ArkaniHamedBaiHeYan, BazierMatteDouvilleMousavandThomasYildirim} describing all polytopal realizations of the $\b{g}$-vector fans of these brick and $2$-acyclic gentle bound quivers as sections of a positive orthant by well-chosen affine spaces (see \cref{subsec:typeConeCA}). Again, this result mainly relies on finding which relations minimally generate the relations between the $\b{g}$-vectors in the non-kissing complex. We present both a purely combinatorial proof (see \cref{subsec:typeConeCA}) and a purely algebraic proof for (non-necessarily gentle) brick algebras making use of extriangulated categories (see \cref{sec:extricats}).
Our algebraic proof is based on an analogue of a result of M.~Auslander~\cite{Auslander1984} relating to minimal generators of Grothendieck groups.
Along the way, we need to generalize, for extriangulated categories, several results from cluster-tilting theory on indices~\cite{DehyKeller} and on abelian quotients~\cite{BuanMarshReiten, KellerReiten} that were known for triangulated categories. We note, however, that our assumptions are not quite the expected ones. This is related to the fact that the Grothendieck groups of cluster categories are badly behaved (\eg the Grothendieck group of a cluster category of type~$A_2$ is trivial). So as to overcome this difficulty, we replace the triangulated structure of cluster categories by some relative extriangulated structure, before considering Grothendieck groups. This explains why we consider projective objects rather than cluster-tilting objects for extriangulated categories.

@@

\medskip
To conclude, it is our hope that the present paper will participate to the interactions between combinatorial geometry, representation theory, and mathematical physics; in particular through further developments of the geometric approach to scattering amplitudes that has flourished during the last years~\cite{ArkaniHamedTrnka-Amplituhedron,ArkaniHamedBourjailyCachazoGoncharovPostnikovTrnka-GrassmannianGeometryScatteringAmplitudes}. The kinematic associahedron from~\cite{ArkaniHamedBaiHeYan} is one of the most recent ``positive geometries''~\cite{ArkaniHamedBaiLam-PositiveGeometries} that has emerged in the study of scattering amplitudes.

@@

\cref{prop::KRextricat} below extends similar results from~\cite{BuanMarshReiten,KellerReiten,KoenigZhu,IyamaYoshino} (see also \cref{prop::functor-F}) to the setting under consideration. We note that the proof requires minor modifications.
